Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-07-19 paper/paperv3.qmd
other phen changes

seal_phenology <- sealdat %>% 
  filter(year >= 2011, 
         moltdur >= 7) %>% 
  mutate(BA = firstobsbreeddoy, 
         BD = firstobsbreeddoy + breeddur, 
         MA = lastobsmoltdoy - moltdur, 
         MD = lastobsmoltdoy) %>% 
  drop_na(age, BA, BD, MA, MD)

@@

pheno_pred <- pheno_grid %>% 
  mutate(predicted = predict(pheno_model, 
                             newdata = pheno_grid,
                             re.form = ~ (1 | year_fct)),
         predicted_se = suppressWarnings(
           # warning: "se.fit computation uses an approximation to estimate the sampling distribution of the parameters"
           predict(pheno_model, 
                   newdata = pheno_grid,
                   re.form = NA,
                   se.fit = TRUE),
         )$se.fit,
         predicted_pop = predict(pheno_model, 
                                 newdata = pheno_grid,
                                 re.form = NA),
         conf_lo = predicted - 1.96 * predicted_se,
         conf_hi = predicted + 1.96 * predicted_se) %>% 
  left_join(n_phen_year, by = "year_fct") %>% 
  group_by(age, age_cat) %>% 
  summarize(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                   \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  # invert logit
  mutate(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                family(pheno_model)$linkinv)) %>% 
  left_join(n_phen_age, by = "age")

phen_pred_years <- ggpredict(
  pheno_model,
  terms = c("age10 [all]", "age_cat", "phase", "year_fct [all]"),
  type = "random",
  interval = "confidence"
) %>% 
  as_tibble() %>% 
  mutate(age = x * 10 + age_senesce,
         age_cat = group,
         phase = facet,
         year_fct = panel) %>% 
  filter((age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce))

breed_pheno_raw <- pheno_raw[pheno_raw$phase == "breeddur",]
molt_pheno_raw <- pheno_raw[pheno_raw$phase == "moltdur",]

breed_duration_plot <- ggpredict(
  pheno_model, 
  terms = c("age10 [all]", "age_cat", "phase", "year_fct [all]")
) %>% 
  as_tibble() %>%
  transmute(age = x * 10 + age_senesce,
            age_cat = group,
            dur_days = predicted,
            dur_days_lwr = conf.low,
            dur_days_upr = conf.high,
            phase = facet,
            year_fct = panel) %>% 
  left_join(n_phen_year, by = "year_fct") %>% 
  filter(phase == "breeddur",
         (age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce)) %>% 
  group_by(age, age_cat) %>% 
  summarize(across(starts_with("dur_days"), \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  ggplot(aes(age)) +
  # Predictions by year
  geom_line(aes(age, predicted, group = interaction(phase, age_cat, year_fct)), 
            filter(phen_pred_years, phase == "breeddur"),
            alpha = 0.1) + 
  # Weighted model (CI + mean)
  geom_ribbon(aes(ymin = dur_days_lwr, ymax = dur_days_upr, fill = age_cat),
              alpha = 0.25) +
  geom_line(aes(y = dur_days, color = age_cat, linetype = age_cat)) +
  # Summarized data
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  breed_pheno_raw,
                  position = position_dodge(width = 0.5)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "Maternal age (years)",
       y = "Breeding haul out (days)") +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 8))

molt_duration_plot <- ggpredict(
  pheno_model, 
  terms = c("age10 [all]", "age_cat", "phase", "year_fct [all]")
) %>% 
  as_tibble() %>%
  transmute(age = x * 10 + age_senesce,
            age_cat = group,
            dur_days = predicted,
            dur_days_lwr = conf.low,
            dur_days_upr = conf.high,
            phase = facet,
            year_fct = panel) %>% 
  left_join(n_phen_year, by = "year_fct") %>% 
  filter(phase == "moltdur",
         (age_cat == "Young" & age < age_senesce) |
           (age_cat == "Old" & age >= age_senesce)) %>% 
  group_by(age, age_cat) %>% 
  summarize(across(starts_with("dur_days"), \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  ggplot(aes(age)) +
  # Predictions by year
  geom_line(aes(age, predicted, group = interaction(phase, age_cat, year_fct)), 
            filter(phen_pred_years, phase == "moltdur"),
            alpha = 0.1) + 
  # Weighted model (CI + mean)
  geom_ribbon(aes(ymin = dur_days_lwr, ymax = dur_days_upr, fill = age_cat),
              alpha = 0.25) +
  geom_line(aes(y = dur_days, color = age_cat, linetype = age_cat)) +
  # Summarized data
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  molt_pheno_raw,
                  position = position_dodge(width = 0.5)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "Maternal age (years)",
       y = "Molt haul out (days)") +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 8))

@@

```{r}
#| label: fig-pheno
#| fig-cap: "Molting gets shorter but breeding doesn't. A: Observed annual time allocation averaged over individuals. B: Raw data for breeding and molting durations (mean ± se) and predictions (mean and 95% CI) from negative binomial weighted by number of seals obersved per year."
#| fig-width: 9
#| fig-height: 9

@@

phen_dist <- seal_phenology %>% 
  # group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "grey50") + 
  labs(x = "Maternal age (Years)", y = "Number of individuals") + 
  theme()

>>>>

seal_phenology <- sealdat %>% 
  filter(year >= 2011, 
         moltdur >= 7, 
         age <= 20) %>% 
  mutate(BA = firstobsbreeddoy, 
         BD = firstobsbreeddoy + breeddur, 
         MA = lastobsmoltdoy - moltdur, 
         MD = lastobsmoltdoy) %>% 
  drop_na(age, BA, BD, MA, MD)

@@

# #population level mean
# phen_plot_pred <- ggpredict(
#   pheno_model, 
#   terms = c("age10 [all]", "age_cat", "phase", "year_fct [all]")
# ) %>% 
#   as_tibble() %>%
#   transmute(age = x * 10 + age_senesce,
#             age_cat = group,
#             dur_days = predicted,
#             dur_days_lwr = conf.low,
#             dur_days_upr = conf.high,
#             phase = facet,
#             year_fct = panel) %>% 
#   left_join(n_phen_year, by = "year_fct") %>% 
#   filter((age_cat == "Young" & age < age_senesce) |
#            (age_cat == "Old" & age >= age_senesce))

#weighted model
pheno_pred <- pheno_grid %>% 
  mutate(predicted = predict(pheno_model, 
                             newdata = pheno_grid,
                             re.form = ~ (1 | year_fct)),
         predicted_se = suppressWarnings(
           # warning: "se.fit computation uses an approximation to estimate the sampling distribution of the parameters"
           predict(pheno_model, 
                   newdata = pheno_grid,
                   re.form = NA,
                   se.fit = TRUE),
         )$se.fit,
         predicted_pop = predict(pheno_model, 
                                 newdata = pheno_grid,
                                 re.form = NA),
         conf_lo = predicted - 1.96 * predicted_se,
         conf_hi = predicted + 1.96 * predicted_se) %>% 
  left_join(n_phen_year, by = "year_fct") %>% 
  group_by(age, age_cat, phase) %>% 
  summarize(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                   \(x) weighted.mean(x, n)),
            .groups = "drop") %>% 
  # invert logit
  mutate(across(c(predicted, predicted_pop, conf_lo, conf_hi), 
                family(pheno_model)$linkinv)) %>% 
  left_join(n_phen_age, by = "age")

#random effect of year
phen_pred_years <- pheno_grid %>% 
  mutate(predicted = predict(pheno_model, 
                             newdata = pheno_grid,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

#raw data
pheno_raw <- seal_phenology %>% 
  select(breeddur, moltdur, age) %>% 
  pivot_longer(c(breeddur, moltdur), 
               names_to = "phase", 
               values_to = "dur_days") %>% 
  group_by(age, phase) %>% 
  summarize(n = n(),
            dur_days_se = sd(dur_days) / sqrt(n),
            dur_days = mean(dur_days),
            dur_days_lwr = dur_days - dur_days_se,
            dur_days_upr = dur_days + dur_days_se,
            .groups = "drop")

breed_duration_plot <- pheno_pred %>% 
  filter(phase == "breeddur") %>% 
  ggplot(aes(age)) +
  # Predictions by year
  geom_line(aes(age, predicted, group = interaction(phase, age_cat, year_fct)), 
            filter(phen_pred_years, phase == "breeddur"),
            alpha = 0.1) + 
  # Weighted model (CI + mean)
  geom_ribbon(aes(ymin = conf_lo, ymax = conf_hi, fill = age_cat),
              alpha = 0.25) +
  geom_line(aes(y = predicted, color = age_cat, linetype = age_cat)) +
  # Summarized data
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  filter(pheno_raw, phase == "breeddur"),
                  position = position_dodge(width = 0.5)) +
    # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  geom_text(aes(label = n), y = 36.5, vjust = 0) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_y_continuous(limits = c(15, 37)) +
  labs(x = "Maternal age (years)",
       y = "Breeding haul out (days)") +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 8))

molt_duration_plot <- pheno_pred %>% 
  filter(phase == "moltdur") %>% 
  ggplot(aes(age)) +
  # Predictions by year
  geom_line(aes(age, predicted, group = interaction(phase, age_cat, year_fct)), 
            filter(phen_pred_years, phase == "moltdur"),
            alpha = 0.1) + 
  # Weighted model (CI + mean)
  geom_ribbon(aes(ymin = conf_lo, ymax = conf_hi, fill = age_cat),
              alpha = 0.25) +
  geom_line(aes(y = predicted, color = age_cat, linetype = age_cat)) +
  # Summarized data
  geom_pointrange(aes(y = dur_days, 
                      ymin = dur_days_lwr, 
                      ymax = dur_days_upr),
                  filter(pheno_raw, phase == "moltdur"),
                  position = position_dodge(width = 0.5)) +
    # Population-level mean response
  geom_line(aes(y = predicted_pop, group = age_cat),
            linetype = "dotted", linewidth = 1.2, alpha = 0.5) +
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  geom_text(aes(label = n), y = 36.5, vjust = 0) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_color_manual(values = c("#7fbc41", "#de77ae")) +
  scale_fill_manual(values = c("#7fbc41", "#de77ae")) +
  scale_y_continuous(limits = c(15, 37)) +
  labs(x = "Maternal age (years)",
       y = "Molt haul out (days)") +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 8))

@@

```{r}
#| label: fig-pheno
#| fig-cap: "Molting gets shorter but breeding doesn't. A: Observed annual time allocation averaged over individuals. B: Raw data for breeding and molting durations (mean ± se) and predictions (mean and 95% CI) from negative binomial weighted by number of seals obersved per year."
#| fig-width: 12
#| fig-height: 9

@@

phen_dist <- seal_phenology %>% 
  # group_by(age) %>% 
  count(age) %>% 
  ggplot(aes(x = age, y = n)) + 
  geom_col(fill = "grey50") + 
  labs(x = "Maternal age (years)", y = "Number of individuals") + 
  theme()
